---
title: "microRNA differential expression analysis"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

Load prerequisites that are needed for differential expression analysis (DEA) in R environment.
```{r, message=FALSE, warning=FALSE}

# libraries
library(tidyverse)
library(edgeR)
library(limma)

```

Read microRNA count data and respective meta information. The raw data (fastqs) and the counts, which are used in this analysis are deployed at GEO under [GSE169569](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE169569) and [GSE169570](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE169570) accession numbers. The data is QCed in terms of outlier samples, the ouliers were removed based on IQR of expressed microRNAs in logarithmic scale (IQR < 1.5).
```{r, warning=FALSE, message=FALSE}

# set wd to main
setwd('..')

# count data
counts <- list.files(
  path = paste0(getwd(), "/data"),
  pattern = "counts.tsv",
  full.names = TRUE
  )
names(counts) <- str_extract(counts, "[a-z]{2}_mirna")

counts <- lapply(counts, function(x){
  read_tsv(x) %>% column_to_rownames("mirna")
})

# meta data
metas <- list.files(
  path = paste0(getwd(), "/data"),
  pattern = "mirna_meta.tsv",
  full.names = TRUE
  )
names(metas) <- str_extract(metas, "[a-z]{2}_mirna")

metas <- lapply(metas, function(x){
  i <- read.table(x, header = TRUE)
  rownames(i) <- i$sample_id
  return(i)
})

```

Summary of the traits in the datasets, where "de" - German (treatment-receiving) and "se" - Swedish (treatment-naive) microRNA datasets:
```{r, warning=FALSE, message=FALSE}

# summarize samples per trait
metas %>%
  bind_rows(.id = "dataset") %>% 
  group_by(dataset, diagnosis) %>% tally() %>%
  bind_rows(., tibble(diagnosis = "TOTAL", n = sum(.$n)))

```

microRNA QC: filtering threshold is average raw counts > 1 in dataset.
```{r, warning=FALSE, message=FALSE}

# summarize mirna expression
qced_mir <- lapply(counts, function(x){
  x %>% as_tibble(rownames = "mir") %>% 
    gather(sample_id, value, -mir) %>% 
    group_by(mir) %>%
    summarise(mean = mean(value)) %>% 
    filter(mean>1) %>% pull(mir)
})

# number of qced mirnas
lapply(qced_mir, length) 
```

Differential expression analysis of microRNA using limma/voom linear models, which include age as a covariate.
```{r, warning=FALSE, message=FALSE}

# generate edgeR objects
dge_list <- vector("list", length = length(counts))
names(dge_list) <- names(counts)

for(set in names(dge_list)){
  dge_list[[set]] <- DGEList(
    counts = counts[[set]][qced_mir[[set]],],
    samples = metas[[set]],
    group = metas[[set]]$diagnosis
  )
}

# calculate size factors
dge_list <- lapply(
  dge_list,
  calcNormFactors
  )
  
# model design
design_list <- lapply(dge_list, function(x){
  model.matrix(
    ~0+diagnosis+age,
    data = x$samples
  )
})

# normalize using voom
dge_list_norm <- lapply(
  1:length(dge_list),
  function(i){
    voomWithQualityWeights(
      dge_list[[i]],
      design_list[[i]],
      plot=FALSE
      )
  })
names(dge_list_norm) <- names(dge_list)

```

Results of treatment-naive Swedish (se) dataset (Supplementary Table 3, 2nd sheet).
```{r, warning=FALSE, message=FALSE}

# contrast matrix
cnts_matrix_se <- makeContrasts(
  "CD_HC"=diagnosisCD-diagnosisHC,
  "UC_HC"=diagnosisUC-diagnosisHC,
  "CD_UC"=diagnosisCD-diagnosisUC,
  "SC_HC"=diagnosisSC-diagnosisHC,
  "CD_SC"=diagnosisCD-diagnosisSC,
  "UC_SC"=diagnosisUC-diagnosisSC,
  levels=colnames(design_list[["se_mirna"]])
)

# fit to a linear model
se_fit <- lmFit(dge_list_norm[["se_mirna"]], design_list[["se_mirna"]])
se_contr_fit <- eBayes(contrasts.fit(se_fit, cnts_matrix_se))

# get results
results_se <- lapply(seq(colnames(cnts_matrix_se)), function(i){
  topTable(se_contr_fit,
           coef=colnames(cnts_matrix_se)[i],
           number=nrow(se_contr_fit),
           sort.by="logFC", confint=T) %>%
    rownames_to_column("Symbol") %>% 
    mutate(
      comparison=gsub("_", " vs ", colnames(cnts_matrix_se)[i]),
      dea = case_when(
        adj.P.Val<0.05 & abs(logFC)>log2(1.5) ~ "de",
        TRUE ~ "non-de")
      )
}) %>% bind_rows()

# write results
setwd('..')
write_tsv(
  results_se,
  paste0(getwd(), "/results/se_mirna_dea_results.tsv")
  )

# print results
head(results_se)

```

Results of treatment-receiving German (de) dataset (Supplementary Table 3, 3rd sheet).
```{r, warning=FALSE, message=FALSE}

# contrast matrix
cnts_matrix_de <- makeContrasts(
  "CD_HC"=diagnosisCD-diagnosisHC,
  "UC_HC"=diagnosisUC-diagnosisHC,
  "CD_UC"=diagnosisCD-diagnosisUC,
  levels=colnames(design_list[["de_mirna"]])
)

# fit to a linear model
de_fit <- lmFit(dge_list_norm[["de_mirna"]], design_list[["de_mirna"]])
de_contr_fit <- eBayes(contrasts.fit(de_fit, cnts_matrix_de))

# get results
results_de <- lapply(seq(colnames(cnts_matrix_de)), function(i){
  topTable(de_contr_fit,
           coef=colnames(cnts_matrix_de)[i],
           number=nrow(de_contr_fit),
           sort.by="logFC", confint=T) %>%
    rownames_to_column("Symbol") %>% 
    mutate(
      comparison=gsub("_", " vs ", colnames(cnts_matrix_de)[i]),
      dea = case_when(
        adj.P.Val<0.05 & abs(logFC)>log2(1.5) ~ "de",
        TRUE ~ "non-de")
      )
}) %>% bind_rows()

# write results
setwd('..')
write_tsv(
  results_de,
  paste0(getwd(), "/results/de_mirna_dea_results.tsv")
  )

# print results
head(results_de)

```

```{r, warning=FALSE, message=FALSE}
sessionInfo()
```


