---
title: "Weighted gene correlation analysis of BeadChip microarray data"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

Load prerequisites that will be needed for weighted gene co-expression network analysis (WGCNA) in R environment.
```{r, message=FALSE, warning=FALSE}

# libraries
library(tidyverse)
library(genefilter)
library(Biobase)
library(WGCNA)

```

Read normalized gene expression intensity values, which were generated using microarray_dea.Rmd script.
```{r, warning=FALSE, message=FALSE}

# set wd to main
setwd('..')

# exprs data
exprs_rna <- read_tsv(
  paste0(getwd(), "/data/se_rna_exprs.tsv")
) %>% column_to_rownames("gene")


# meta data
meta_rna <- read_tsv(
  paste0(getwd(),"/data/se_rna_meta.tsv")
  )

meta_mirna <- read_tsv(
  paste0(getwd(),"/data/se_mirna_meta.tsv")
  )

# keep overlapping
meta_rna <- meta_rna %>% 
  filter(sample_id %in% meta_mirna$sample_id)

```

Retain the most variable genes using varFilter function from genefilter.
```{r, warning=FALSE, message=FALSE}

# keep variable genes
exprs_rna_filt <- exprs(
  varFilter(
    ExpressionSet(
      assayData = as.matrix(exprs_rna)
      ),
    var.func = "sd",
    var.cutoff=0.5
    )
  )
# print number of genes
dim(exprs_rna_filt)

```

Filter outlying samples based on network connectivity (WGCNA workflow):
```{r, warning=FALSE, message=FALSE}

# split expression per trait
exprs_list <- exprs_rna_filt %>% t() %>%
  as_tibble(rownames = "array_id") %>% 
  left_join(
    meta_rna %>%
      select(array_id, diagnosis),., by = "array_id"
    ) %>%
  column_to_rownames("array_id") %>%
  split(., .$diagnosis) %>% 
  lapply(., function(x){
    x %>% select(-diagnosis) %>% t()
    })

# squared Euclidean distance
adj_list <- lapply(
  exprs_list,
  adjacency,
  type = "distance"
  )

# connectivity
k_list <- lapply(adj_list, function(x){
  as.numeric(apply(x, 2, sum)) - 1
  })

# standardized connectivity
Z_k_list <- lapply(k_list, scale)

# threshold
thresholdZ_k <- -2.5

# get outlying samples
remove_list <- 
  lapply(Z_k_list, function(x){
    x < thresholdZ_k | is.na(x)
    })

# remove the outliers from expression data
exprs_list_qced <- 
  lapply(seq(exprs_list), function(x) {
    exprs_list[[x]][,!remove_list[[x]]]
    })
names(exprs_list_qced) <- names(exprs_list)

# print dims
lapply(exprs_list_qced, dim)

```

Calculate soft-thresholding powers for construction of trait-wise networks.
```{r, warning=FALSE, message=FALSE}

# set of soft thresholding powers
powers = c(1:20)

# power based on SFT criterion
sft_list <- lapply(exprs_list_qced, function(x){
  pickSoftThreshold(
    t(x),
    powerVector = powers,
    networkType = "signed",
    RsquaredCut = 0.8
    )
})

# add signed value of SFT R^2
sft_list <- lapply(sft_list, function(x) {
  x$fitIndices <- x$fitIndices %>% 
    mutate(SFT.R.sq.signed = -sign(slope)*SFT.R.sq)
  return(x)
  })

```

Calculate topological overlap matrices for each trait
```{r, warning=FALSE, message=FALSE}

# get median threshold of networks
sft <- lapply(
  sft_list, function(x) x$powerEstimate
  ) %>% 
  bind_rows() %>%
  gather(diagnosis, soft) %>%
  summarise(soft=round(median(soft))) %>% 
  pull(soft)

# list of topological overlap matrices
toms_list <- lapply(exprs_list_qced, function(x){
  TOMsimilarityFromExpr(
    t(x),
    corType = "bicor",
    networkType = "signed",
    power = sft,
    TOMDenom = "mean",
    nThreads = 4
    )
  })

# add gene names
toms_list <- lapply(toms_list, function(x){
  colnames(x) <- rownames(exprs_list_qced$HC)
  rownames(x) <- rownames(exprs_list_qced$HC)
  return(x)
})

# write for tensor decomposition
setwd('..')
for(i in seq(toms_list)){
  x <- toms_list[[i]] %>% as.data.frame()
  write_csv(
    x, path = gzfile(
      paste0(getwd(), "/data/tom_mrna_", names(toms_list)[i], ".csv.gz")
      )
  )
}

# print dims
lapply(toms_list, dim)

```

```{r, warning=FALSE, message=FALSE}
sessionInfo()
```
